Scaling rates of true polar wander in convecting planets and moons

### 2017-09-13 tpw_rate.tex
Fix typo.

In the limit that the eigengap becomes zero, a small perturbation $\delta$ can produce an arbitrary rotation of the principal axes. In effect, the location of the principal axes are set by the location of a small density anomaly. This has a close correspondence with hypothesized IITPW \citep{kirschvink1997evidence}. As noted above, the interchange of the maximum and intermediate axex can produce a large change in $\theta$ because the misalignment angle is suddenly measured from a new axis. The resulting TPW event can produce a large change in the orientation of the rotation axis (potentially $90^{\circ}$ if there are not further perturbations to the $\mathbf{E}$ tensor).  However, the eigengap does not need to be zero for there to be large polar wander events. In addition, wander does not need to be $90^\circ$ after an interchange of maximum and intermediate axes due to the subsequent evolution of density anomalies during the event.  We clarify these points using numerical simulations of convection in a 2D spherical annulus.

@@

We can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch} to obtain an estimate of maximum TPW rates for Earth.
Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
\citet{richards1997explanation} performed TPW simulations based on Cenozoic plate motion reconstructions, and argued that the convective planform of Earth has been stable for the last few hundred million years. 
\citet{cambiotti2011new} took a similar approach, but allowed for the position of the spin axis to lag the principal axes (i.e. nonzero $\theta$). In that study they found values for $\theta$ of up to 3$^\circ$-7$^\circ$.
On the other hand, we know that there have been large reorganizations of that planform during Earth history,
so this relative stability of the spin-axis may not hold in general \citep{evans2003true}.
Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
\begin{equation}
\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{\tau_R} \sim 6^\circ / \mathrm{Myr},
\end{equation}
which corresponds to about 66 cm/yr at a point 90$^\circ$ from the TPW axis (and is comparable to the maximum rates seen in Figure~\ref{fig:misfit}.
This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the
possibility of a larger mismatch angle. Such large angles may be necessary to explain the range of rates suggested by some interpretations of paleomagnetic data \citep{mitchell2011sutton}.
The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
which results in a corresponding uncertainty for the relaxation time $\tau_R$ and the maximum polar wander rate.

>>>>

In the limit that the eigengap becomes zero, a small perturbation $\delta$ can produce an arbitrary rotation of the principal axes. In effect, the location of the principal axes are set by the location of a small density anomaly. This has a close correspondence with hypothesized IITPW \citep{kirschvink1997evidence}. As noted above, the interchange of the maximum and intermediate axes can produce a large change in $\theta$ because the misalignment angle is suddenly measured from a new axis. The resulting TPW event can produce a large change in the orientation of the rotation axis (potentially $90^{\circ}$ if there are not further perturbations to the $\mathbf{E}$ tensor).  However, the eigengap does not need to be zero for there to be large polar wander events. In addition, wander does not need to be $90^\circ$ after an interchange of maximum and intermediate axes due to the subsequent evolution of density anomalies during the event.  We clarify these points using numerical simulations of convection in a 2D spherical annulus.

@@

We can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch} to obtain an estimate of maximum TPW rates for Earth.
Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
\citet{richards1997explanation} performed TPW simulations based on Cenozoic plate motion reconstructions, and argued that the convective planform of Earth has been stable for the last few hundred million years. 
\citet{cambiotti2011new} took a similar approach, but allowed for the position of the spin axis to lag the principal axes (i.e. nonzero $\theta$). In that study they found values for $\theta$ of up to 3$^\circ$-7$^\circ$.
On the other hand, we know that there have been large reorganizations of that planform during Earth history,
so this relative stability of the spin-axis may not hold in general \citep{evans2003true}.
Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
\begin{equation}
\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{\tau_R} \sim 6^\circ / \mathrm{Myr},
\end{equation}
which corresponds to about 66 cm/yr at a point 90$^\circ$ from the TPW axis (and is comparable to the maximum rates seen in Figure~\ref{fig:misfit}).
This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the
possibility of a larger mismatch angle. Such large angles may be necessary to explain the range of rates suggested by some interpretations of paleomagnetic data \citep{mitchell2011sutton}.
The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
which results in a corresponding uncertainty for the relaxation time $\tau_R$ and the maximum polar wander rate.
